Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 response-to-reviewers.tex

% LaTeX rebuttal letter example.
%
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
%
% Licensed under cc by-sa 3.0 with attribution required.

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule}
                  % \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent
    {\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication.  We have addressed
the points you and Reviewer 1 raised below in turn.

\reviewersection
\section*{Reviewer 1 comments}

\begin{point}
The authors present VCF Zarr, a specification that translates the variant call
format (VCF) data model into an array-based representation for the Zarr storage
format. They also present the `vcf2zarr` utility to convert large VCFs to Zarr.
They provide data compression and analysis benchmarks comparing VCF Zarr to
existing variant storage technologies using simulated genotype data. They also
present a case study on real world Genomics England aggV2 data.

The authors' benchmarks overall show that VCF Zarr has superior compression and
computational analysis performance at scale relative to data stored as
row-oriented VCF and that VCF Zarr is competitive with specialized storage
solutions that require similarly specialized tools and access libraries for
querying. An attractive feature is that VCF Zarr allows for variant annotation
workflows that do not require full dataset copy and conversion. Another key
point is that Zarr is a high-level spec and data model for the chunked storage
of n-d arrays, rather than a byte-level encoding designed specifically around
the genomic variant data type. I personally have used Zarr productively for
several applications unrelated to statistical genetics. While Zarr VCF mildly
underperforms some of the specialized formats (Savvy in compute, Genozip in
compression) in a few instances, I believe the accessibility, interoperability,
and reusability gains of Zarr make the small tradeoff well worthwhile.

Because Zarr has seen heavy adoption in other scientific communities like the
geospatial and Earth sciences, and is well integrated in the scientific Python
stack, I think it holds potential for greater reusability across the ecosystem.
As such, I think the VCF Zarr spec is a highly valuable if not overdue
contribution to an entrenched field that has recently been confronted by a
scalability wall.

Overall, the paper is clear, comprehensive, and well written. 
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The benefits for large scientific datasets to be analysis-ready
cloud-optimized (ARCO) have been well articulated by Abernathey et al., 2021.
However, I do think that the "local"/HPC single-file use case is still
important and won't disappear any time soon, and for some file system use
cases, expansive and deep hierarchies can be performance limiting (this was
hinted at in one of the benchmarks). In this scenario would a large Zarr VCF
perform reasonably well (or even better on some file systems) via a single
local zip store?
\end{point}
\begin{reply}
\end{reply}

\begin{point}
The description of the intermediate columnar format (ICF) used by `vcf2zarr`
is missing some detail. At first I got the impression it might be based on
something like Parquet, but running the provided code showed that it consists
of a similar file-based chunk layout to Zarr. This should be clarified in the
manuscript.

\end{point}
\begin{reply}
\end{reply}

\begin{point}
The authors discuss the possibility of storing an index mapping genomic
coordinates to chunk indexes. Have Zarr-based formats in other fields like
geospatial introduced their own indexing approaches to take inspiration from?
\end{point}
\begin{reply}
\end{reply}

\begin{point}
Since VCF Zarr is still a draft proposal, it could be useful to indicate
where community discussions are happening and how potential new contributors
can get involved, if possible. This doesn't need to be in the paper per se, but
perhaps documented in the spec repo.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
In the background: "For the representation to be FAIR, it must also be
accessible," -- A is for "accessible", so "also" doesn't make sense.
\end{point}
\begin{reply}
\end{reply}

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
\end{reply}

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
\end{reply}

\reviewersection
\section*{Reviewer 2 comments}

\begin{point}
Point
\end{point}
\begin{reply}
\end{reply}

\end{document}

### 2025-01-30 paper.tex
Add GSC ref
Nothing much to discuss, it's another method like genozip that just
offers VCF output format (well, supports plink also).

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression
and query/compute performance are well
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries,
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
Several specialised algorithms for compressing
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
(See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.)
The widely-used PLINK binary format stores genotypes in a
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{rivas2024efficient} is more general
and compact than PLINK, compressing variant data using specialised
algorithms~\cite{sambo2014compression}.
Methods have also been developed which store variation data
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.
The SeqArray package~\citep{zheng2017seqarray} builds on the
Genomic Data Storage container format~\cite{zheng2012high}
to store VCF genotype data in a packed and compressed format,
and is used in several downstream R packages~\cite[e.g.][]{
gogarten2019genetic,fernandes2020simplephenotypes}.

@@

The discrepancy between elapsed time and total CPU time for Zarr
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
of accessing a large number of small chunk files and
consequent reduced effectiveness of operating system read-ahead caching.
While having many independent chunks maps
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting}
(see also the section on accessing Genomics England data stored on S3),
such file access patterns can be problematic in the
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory
hierarchy) provides a useful alternative here and can
significantly reduce overall processing time. To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75 GiB and 24.24 GiB
for file system and Zip file, respectively)
and while total
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252

>>>>

\subsection{Storing genetic variation data}
Although VCF is the standard format for exchanging genetic variation
data, its limitations both in terms of compression
and query/compute performance are well
known~\citep[e.g.][]{kelleher2013processing,layer2016efficient,li2016bgt},
and many methods
have been suggested to improve on these properties.
Most approaches balance compression with
performance on particular types of queries,
typically using a command line interface (CLI)
and outputting VCF text~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc,%GBC
luo2024gsc}. %GSC
Several specialised algorithms for compressing
the genotype matrix (i.e., just the genotype calls without additional
VCF information) have been proposed
\citep{qiao2012handling, % SpeedGene
deorowicz2013genome, %TGC
sambo2014compression, % snpack
deorowicz2019gtshark, %GTShark
deorowicz2021vcfshark, % VCFShark
dehaas2024genotype} %  GRG
most notably the Positional
Burrows--Wheeler Transform (PBWT)~\citep{durbin2014efficient}.
(See~\citep{mcvean2019linkage} for a review of the techniques
employed in genetic data compression.)
The widely-used PLINK binary format stores genotypes in a
packed binary representation, supporting only biallelic
variants without phase information.
The PLINK 2 PGEN format~\citep{rivas2024efficient} is more general
and compact than PLINK, compressing variant data using specialised
algorithms~\cite{sambo2014compression}.
Methods have also been developed which store variation data
along with annotations in databases to facilitate
efficient queries~\cite[e.g.][]{
paila2013gemini,%GEMINI
lopez2017hgva} %OpenCGA
which either limit to certain classes of variant~\cite[e.g.][]{greene2023genetic}
or have storage requirements larger
than uncompressed VCF~\citep{al2023critical}.
The SeqArray package~\citep{zheng2017seqarray} builds on the
Genomic Data Storage container format~\cite{zheng2012high}
to store VCF genotype data in a packed and compressed format,
and is used in several downstream R packages~\cite[e.g.][]{
gogarten2019genetic,fernandes2020simplephenotypes}.

@@

The discrepancy between elapsed time and total CPU time for Zarr
benchmarks in
Figs~\ref{fig-whole-matrix-compute}--~\ref{fig-column-extract}
and~\ref{fig-whole-matrix-decode} is mostly due to the overhead
of accessing a large number of small chunk files and
consequent reduced effectiveness of operating system read-ahead caching.
While having many independent chunks maps
naturally to high-performance parallel access patterns
in the cloud store setting~\citep{durner2023exploiting}
(see also the section on accessing Genomics England data stored on S3),
such file access patterns can be problematic in the
classical file-system setting.
The Zarr Zip store (which is simply a Zip archive of the standard directory
hierarchy) provides a useful alternative here and can
significantly reduce overall processing time. 
(The nascent Zarr v3 specification provides another option 
via  ``sharding'' which essentially 
allows multiple chunks to be stored in a single file.)
To illustrate this
we repeated the benchmarks in Fig~\ref{fig-whole-matrix-decode} in
which we measure the time required to decode the entire genotype
matrix (thus serving as a lower-bound for any subsequent computations).
In the case of $10^6$ samples, the Zip file requires slightly more
% Full data table above ^
% 58         48129895      1000000    7254858                zarr    23.752762
% 59         48129895      1000000    7254858            zarr.zip    24.240338
storage space than the directory hierarchy (23.75 GiB and 24.24 GiB
for file system and Zip file, respectively)
and while total
CPU time is similar (193m vs 187m), the elapsed wall clock
time for the Zip file version is significantly less
(258m vs 188m).
        % tool  user_time  sys_time    cpu_time   wall_time
% 11      zarr   11383.82    191.01  192.913833  258.598778
% 29  zarr.zip   11198.37     43.13  187.358333  188.659252
